### 2010-09-10 latex/dissertacao/abstract.tex
Levantei a bola do BCR+ e fiz outras pequenas alteracoes.

This work presents three dependency models in order to support the evaluation of reverse engineering tools via controlled simulations. An evaluation of the models showed that, with an appropriate choice of parameters, they produce dependency networks that are structurally similar to dependencies extracted from real software systems. This work also presents a proof of concept, demonstrating the feasibility of using one of the models to evaluate algorithms used in the context of software architecture recovery, a branch of reverse engineering.

>>>>

This paper presents computer models that generate representations which can be interpreted as dependencies between source code entities. One of the models, called BCR+, was developed in the context of this work. We have also developed a classification model that can tell, with accuracy of 96\%, whether a representation of dependencies is realistic --- that is, if it resembles representations extracted from real systems. %Then, we identified parameter settings for three computational models (among them, BCR+) that favor the generation of realistic representations of dependencies.

% This work presents three dependency models in order to support the evaluation of reverse engineering tools via controlled simulations. One of them, called BCR+, was developed in the context of this work. An evaluation of the models showed that, with an appropriate choice of parameters, they produce dependency networks that are structurally similar to dependencies extracted from real software systems. Moreover, classification rules were derived to predict, with 75\% accuracy, whether a network that was generated with certain parameter values will be similar to networks extracted from software systems.

This work also presents a proof of concept, demonstrating the feasibility of using BCR+ to evaluate algorithms used in the context of software architecture recovery, a branch of reverse engineering.

### 2010-09-10 latex/dissertacao/apendices.tex
Levantei a bola do BCR+ e fiz outras pequenas alteracoes.

\begin{chapter}{Redes Empregadas na Avaliação dos Modelos}
% select nme_network, to_char(s_score, '0.99'), n_vertices, n_edges 
% from view_network
% where fk_classification = 1
% OR nme_network ilike '%irpf%'
%
\begin{center}
\begin{longtable}{| p{10cm} | c | r | r |}
	\hline
	\textbf{Sistema} & \textbf{Valor} S & \textbf{Vértices} & \textbf{Arestas} \\ \hline
	\hline
	AbaGuiBuilder-1.8 &  $0,97$ & 2929 & 12399 \\ \hline
	alfresco-labs-deployment-3Stable &  $0,97$ & 2441 & 9178 \\ \hline
	aoi272 &  $0,98$ & 1579 & 11670 \\ \hline
	ArgoUML-0.28 &  $0,97$ & 5075 & 29018 \\ \hline
	battlefieldjava-0.1 &  $0,98$ & 707 & 2038 \\ \hline
	broker-4.1.5 &  $0,97$ & 5940 & 25589 \\ \hline
	checkstyle-5.0 &  $0,98$ & 1174 & 4513 \\ \hline
	dbwrench &  $0,98$ & 2015 & 10188 \\ \hline
	dom4j-1.6.1 &  $0,98$ & 4666 & 31689 \\ \hline
	ec2-api-tools &  $0,98$ & 5255 & 26916 \\ \hline
	ermodeller-1.9.2-binary &  $0,97$ & 1441 & 6261 \\ \hline
	findbugs-1.3.8 &  $0,98$ & 3401 & 20023 \\ \hline
	flyingsaucer-R8 &  $0,97$ & 1323 & 6610 \\ \hline
	freetts-1.2.2-bin &  $0,91$ & 273 & 1106 \\ \hline
	ganttproject-2.0.9 &  $0,98$ & 5447 & 25558 \\ \hline
	gdata-src.java-1.31.1 &  $0,97$ & 1649 & 8150 \\ \hline
	GEF-0.13-bin &  $0,97$ & 352 & 1484 \\ \hline
	geoserver-2.0-beta1-bin &  $0,97$ & 25007 & 145034 \\ \hline
	geotools-2.5.5-bin &  $0,97$ & 13559 & 79499 \\ \hline
	gfp\_0.8.1 &  $0,87$ & 475 & 1030 \\ \hline
	guice-2.0 &  $0,98$ & 638 & 2980 \\ \hline
	gwt-windows-1.6.4 &  $0,98$ & 6065 & 32729 \\ \hline
	hibernate-distribution-3.3.1.GA-dist &  $0,98$ & 4340 & 22410 \\ \hline
	Hl7Comm.1.0.1 &  $0,97$ & 337 & 1335 \\ \hline
	hsqldb\_1\_8\_0\_10 &  $0,95$ & 369 & 1901 \\ \hline
	iBATIS\_DBL-2.1.5.582 &  $0,95$ & 291 & 1300 \\ \hline
	iFreeBudget-2.0.9 &  $0,98$ & 4053 & 20028 \\ \hline
	iReport-nb-3.5.1 &  $0,97$ & 35019 & 184701 \\ \hline
	IRPF2009v1.1 &  $0,98$ & 5396 & 29247 \\ \hline
	JabRef-2.5b2-src &  $0,98$ & 2473 & 10437 \\ \hline
	jai-1\_1\_4-pre-dr-b03-lib-linux-i586 &  $0,96$ & 936 & 5086 \\ \hline
	jailer\_2.9.9 &  $0,97$ & 3691 & 14890 \\ \hline
	jakarta-tomcat-5.0.28-embed &  $0,98$ & 2314 & 10652 \\ \hline
	jalopy-1.5rc3 &  $0,98$ & 888 & 4349 \\ \hline
	jasperreports-3.5.2-project &  $0,98$ & 20508 & 120243 \\ \hline
	jfreechart-1.0.13 &  $0,97$ & 1989 & 13891 \\ \hline
	jGnash-2.2.0 &  $0,98$ & 6948 & 38303 \\ \hline
	jgraphpad-5.10.0.2 &  $0,96$ & 519 & 2257 \\ \hline
	jmsn-0.9.9b2 &  $0,83$ & 232 & 777 \\ \hline
	juel-2.1.2 &  $0,94$ & 111 & 435 \\ \hline
	juxy-0.8 &  $0,97$ & 3025 & 20427 \\ \hline
	JXv3.2rc2deploy &  $0,96$ & 711 & 2573 \\ \hline
	makagiga-3.4 &  $0,98$ & 1999 & 10819 \\ \hline
	MegaMek-v0.34.3 &  $0,98$ & 2855 & 19575 \\ \hline
	mondrian-3.1.1.12687 &  $0,97$ & 2000 & 14128 \\ \hline
	myjgui\_0.6.6 &  $0,96$ & 604 & 2430 \\ \hline
	oddjob-0.26.0 &  $0,98$ & 4204 & 18752 \\ \hline
	openxava-3.1.2 &  $0,98$ & 16919 & 89049 \\ \hline
	pdfsam-1.1.3-out &  $0,97$ & 3043 & 16024 \\ \hline
	peer-4.1.5 &  $0,97$ & 9291 & 50614 \\ \hline
	pentaho-reporting-engine-classic-0.8.9.11 &  $0,98$ & 3454 & 17948 \\ \hline
	pjirc\_2\_2\_1\_bin &  $0,84$ & 180 & 687 \\ \hline
	pmd-bin-4.2.5 &  $0,98$ & 1719 & 8098 \\ \hline
	proguard4.3 &  $0,98$ & 604 & 4923 \\ \hline
	rapidminer-4.4-community &  $0,96$ & 11764 & 55976 \\ \hline
	smc\_6\_0\_0 &  $0,97$ & 698 & 3571 \\ \hline
	squirrel-sql-3.0.1-base &  $0,98$ & 8982 & 43806 \\ \hline
	squirrel-sql-3.0.1-standard &  $0,98$ & 11922 & 53969 \\ \hline
	stendhal-0.74 &  $0,97$ & 747 & 3056 \\ \hline
	subethasmtp-3.1 &  $0,98$ & 1485 & 8174 \\ \hline
	thinkui\_sqlclient-1.1.2 &  $0,97$ & 2883 & 21979 \\ \hline
	tvbrowser-2.7.3-bin &  $0,97$ & 3335 & 14110 \\ \hline
	villonanny-2.3.0.b02.bin &  $0,98$ & 1949 & 6708 \\ \hline
	worker-4.1.5 &  $0,97$ & 5940 & 25589 \\ \hline
	zk-bin-3.6.1 &  $0,95$ & 13911 & 82219 \\ \hline

	\caption{\label{tab:redes}Sistemas de software usados no estudo de software-realismo de modelos e respectivos valores S.}
\end{longtable}
\end{center}

\begin{center}
\begin{longtable}{| p{6cm} | l | r | r | }
	\hline
	\textbf{Rede} & \textbf{Valor} S & \textbf{Vértices} & \textbf{Arestas} \\ \hline
	\hline
	5 redes de amizade do site Facebook \cite{Traud2008} & $-0,130 \pm 0,007$ & 768 a 18162 & 33312 a 1533600 \\ \hline
	3 redes de circuitos eletrônicos \cite{Milo2004} & $0,433 \pm 0,005$ & 121 a 511 & 189 a 819 \\ \hline
	4 redes de adjacências entre palavras \cite{Milo2004} & $0,573 \pm 0,064$ & 2703 a 11585 & 8300 a 46281 \\ \hline
	3 redes de estrutura proteica \cite{Milo2004} & $0,540 \pm 0,046$ & 52 a 96 & 123 a 213 \\ \hline
	2 redes sociais de sentimento positivo \cite{Milo2004} & $0,555 \pm 0,248$ & 31 a 66 & 96 a 182 \\ \hline
	43 redes metabólicas \cite{Jeong2000} & $0,509 \pm 0,015$ & 408 a 2360 & 792 a 5959 \\ \hline
	Rede de interação entre proteínas da levedura \cite{Jeong2001} & $0,22$ & 687 & 1079 \\ \hline
	Links entre blogs políticos \cite{Adamic2005} & $0,96$ & 1489 & 19090 \\ \hline
	Rede neural do verme C Elegans \cite{Watts1998} & $0,87$ & 296 & 2359 \\ \hline
	Rede ``beta3sreduced'' (fonte desconhecida) & $0,05$ & 1286 & 33813 \\ \hline
	Rede ``czech'' (fonte desconhecida) & $0,78$ & 5225 & 51687 \\  \hline
	Rede ``ecoli-metabolic'' (fonte desconhecida) & $0,69$ & 895 & 964 \\ \hline

	\caption{\label{tab:redes-outros}Redes de domínios diversos usados no estudo de software-realismo de modelos e respectivos valores S.}
\end{longtable}
\end{center}

@@

No contexto de recuperação de arquitetura de software, é comum descrever as entidades de código-fonte como um grafo de dependências entre entidades e usar a métrica de Jaccard para medir a similaridade entre duas entidades \cite{Anquetil1999}. Sejam X e Y duas entidades, e seja d(A) o conjunto de entidades ligadas à entidade A no grafo de dependências. A similaridade entre X e Y é dada pela expressão a seguir:

@@

% TODO: fórmula da QM

@@

% TODO: reescrever sem emoção, afinal isto virou um apêndice. Talvez mover o início para o capítulo do estudo.

>>>>

\begin{chapter}{Redes Empregadas na Avaliação dos Modelos} \label{cap:lista-redes}
% select nme_network, to_char(s_score, '0.99'), n_vertices, n_edges 
% from view_network
% where fk_classification = 1
% OR nme_network ilike '%irpf%'
%

\begin{center}
\begin{longtable}{| p{10cm} | c | r | r |}
	\caption{65 redes de software (conjunto $T$).}	\\
	\hline
	\textbf{Sistema} & \textbf{S($x$, $T$)} & \textbf{Vértices} & \textbf{Arestas} \\ \hline
	\hline
	AbaGuiBuilder-1.8 &  $0,97$ & 2929 & 12399 \\ \hline
	alfresco-labs-deployment-3Stable &  $0,97$ & 2441 & 9178 \\ \hline
	aoi272 &  $0,98$ & 1579 & 11670 \\ \hline
	ArgoUML-0.28 &  $0,97$ & 5075 & 29018 \\ \hline
	battlefieldjava-0.1 &  $0,98$ & 707 & 2038 \\ \hline
	broker-4.1.5 &  $0,97$ & 5940 & 25589 \\ \hline
	checkstyle-5.0 &  $0,98$ & 1174 & 4513 \\ \hline
	dbwrench &  $0,98$ & 2015 & 10188 \\ \hline
	dom4j-1.6.1 &  $0,98$ & 4666 & 31689 \\ \hline
	ec2-api-tools &  $0,98$ & 5255 & 26916 \\ \hline
	ermodeller-1.9.2-binary &  $0,97$ & 1441 & 6261 \\ \hline
	findbugs-1.3.8 &  $0,98$ & 3401 & 20023 \\ \hline
	flyingsaucer-R8 &  $0,97$ & 1323 & 6610 \\ \hline
	freetts-1.2.2-bin &  $0,91$ & 273 & 1106 \\ \hline
	ganttproject-2.0.9 &  $0,98$ & 5447 & 25558 \\ \hline
	gdata-src.java-1.31.1 &  $0,97$ & 1649 & 8150 \\ \hline
	GEF-0.13-bin &  $0,97$ & 352 & 1484 \\ \hline
	geoserver-2.0-beta1-bin &  $0,97$ & 25007 & 145034 \\ \hline
	geotools-2.5.5-bin &  $0,97$ & 13559 & 79499 \\ \hline
	gfp\_0.8.1 &  $0,87$ & 475 & 1030 \\ \hline
	guice-2.0 &  $0,98$ & 638 & 2980 \\ \hline
	gwt-windows-1.6.4 &  $0,98$ & 6065 & 32729 \\ \hline
	hibernate-distribution-3.3.1.GA-dist &  $0,98$ & 4340 & 22410 \\ \hline
	Hl7Comm.1.0.1 &  $0,97$ & 337 & 1335 \\ \hline
	hsqldb\_1\_8\_0\_10 &  $0,95$ & 369 & 1901 \\ \hline
	iBATIS\_DBL-2.1.5.582 &  $0,95$ & 291 & 1300 \\ \hline
	iFreeBudget-2.0.9 &  $0,98$ & 4053 & 20028 \\ \hline
	iReport-nb-3.5.1 &  $0,97$ & 35019 & 184701 \\ \hline
	IRPF2009v1.1 &  $0,98$ & 5396 & 29247 \\ \hline
	JabRef-2.5b2-src &  $0,98$ & 2473 & 10437 \\ \hline
	jai-1\_1\_4-pre-dr-b03-lib-linux-i586 &  $0,96$ & 936 & 5086 \\ \hline
	jailer\_2.9.9 &  $0,97$ & 3691 & 14890 \\ \hline
	jakarta-tomcat-5.0.28-embed &  $0,98$ & 2314 & 10652 \\ \hline
	jalopy-1.5rc3 &  $0,98$ & 888 & 4349 \\ \hline
	jasperreports-3.5.2-project &  $0,98$ & 20508 & 120243 \\ \hline
	jfreechart-1.0.13 &  $0,97$ & 1989 & 13891 \\ \hline
	jGnash-2.2.0 &  $0,98$ & 6948 & 38303 \\ \hline
	jgraphpad-5.10.0.2 &  $0,96$ & 519 & 2257 \\ \hline
	jmsn-0.9.9b2 &  $0,83$ & 232 & 777 \\ \hline
	juel-2.1.2 &  $0,94$ & 111 & 435 \\ \hline
	juxy-0.8 &  $0,97$ & 3025 & 20427 \\ \hline
	JXv3.2rc2deploy &  $0,96$ & 711 & 2573 \\ \hline
	makagiga-3.4 &  $0,98$ & 1999 & 10819 \\ \hline
	MegaMek-v0.34.3 &  $0,98$ & 2855 & 19575 \\ \hline
	mondrian-3.1.1.12687 &  $0,97$ & 2000 & 14128 \\ \hline
	myjgui\_0.6.6 &  $0,96$ & 604 & 2430 \\ \hline
	oddjob-0.26.0 &  $0,98$ & 4204 & 18752 \\ \hline
	openxava-3.1.2 &  $0,98$ & 16919 & 89049 \\ \hline
	pdfsam-1.1.3-out &  $0,97$ & 3043 & 16024 \\ \hline
	peer-4.1.5 &  $0,97$ & 9291 & 50614 \\ \hline
	pentaho-reporting-engine-classic-0.8.9.11 &  $0,98$ & 3454 & 17948 \\ \hline
	pjirc\_2\_2\_1\_bin &  $0,84$ & 180 & 687 \\ \hline
	pmd-bin-4.2.5 &  $0,98$ & 1719 & 8098 \\ \hline
	proguard4.3 &  $0,98$ & 604 & 4923 \\ \hline
	rapidminer-4.4-community &  $0,96$ & 11764 & 55976 \\ \hline
	smc\_6\_0\_0 &  $0,97$ & 698 & 3571 \\ \hline
	squirrel-sql-3.0.1-base &  $0,98$ & 8982 & 43806 \\ \hline
	squirrel-sql-3.0.1-standard &  $0,98$ & 11922 & 53969 \\ \hline
	stendhal-0.74 &  $0,97$ & 747 & 3056 \\ \hline
	subethasmtp-3.1 &  $0,98$ & 1485 & 8174 \\ \hline
	thinkui\_sqlclient-1.1.2 &  $0,97$ & 2883 & 21979 \\ \hline
	tvbrowser-2.7.3-bin &  $0,97$ & 3335 & 14110 \\ \hline
	villonanny-2.3.0.b02.bin &  $0,98$ & 1949 & 6708 \\ \hline
	worker-4.1.5 &  $0,97$ & 5940 & 25589 \\ \hline
	zk-bin-3.6.1 &  $0,95$ & 13911 & 82219 \\ \hline

	% \caption{\label{tab:redes}Sistemas de software usados no estudo de software-realismo de modelos e respectivos valores S.}
\end{longtable}
\end{center}

\begin{center}
\begin{longtable}{| p{6cm} | l | r | r | }
	\caption{66 redes de domínios diversos (conjunto $\bar{T}$).} \\
	\hline
	\textbf{Rede} & \textbf{S($x$, $T$)} & \textbf{Vértices} & \textbf{Arestas} \\ \hline
	\hline
	5 redes de amizade do site Facebook \cite{Traud2008} & $-0,130 \pm 0,007$ & 768 a 18162 & 33312 a 1533600 \\ \hline
	3 redes de circuitos eletrônicos \cite{Milo2004} & $0,433 \pm 0,005$ & 121 a 511 & 189 a 819 \\ \hline
	4 redes de adjacências entre palavras \cite{Milo2004} & $0,573 \pm 0,064$ & 2703 a 11585 & 8300 a 46281 \\ \hline
	3 redes de estrutura proteica \cite{Milo2004} & $0,540 \pm 0,046$ & 52 a 96 & 123 a 213 \\ \hline
	2 redes sociais de sentimento positivo \cite{Milo2004} & $0,555 \pm 0,248$ & 31 a 66 & 96 a 182 \\ \hline
	43 redes metabólicas \cite{Jeong2000} & $0,509 \pm 0,015$ & 408 a 2360 & 792 a 5959 \\ \hline
	Rede de interação entre proteínas da levedura \cite{Jeong2001} & $0,22$ & 687 & 1079 \\ \hline
	Links entre blogs políticos \cite{Adamic2005} & $0,96$ & 1489 & 19090 \\ \hline
	Rede neural do verme C Elegans \cite{Watts1998} & $0,87$ & 296 & 2359 \\ \hline
	Rede ``beta3sreduced'' (fonte desconhecida) & $0,05$ & 1286 & 33813 \\ \hline
	Rede ``czech'' (fonte desconhecida) & $0,78$ & 5225 & 51687 \\  \hline
	Rede ``ecoli-metabolic'' (fonte desconhecida) & $0,69$ & 895 & 964 \\ \hline

	% \caption{\label{tab:redes-outros}Redes de domínios diversos usados no estudo de software-realismo de modelos e respectivos valores S.}
\end{longtable}
\end{center}

@@

No contexto de recuperação de arquitetura de software, é comum descrever as entidades de código-fonte como um grafo de dependências entre entidades e usar a métrica de Jaccard para medir a similaridade entre duas entidades \cite{Anquetil1999}. Sejam X e Y duas entidades, e seja d(A) o conjunto de entidades ligadas à entidade A no grafo de dependências. A similaridade entre duas entidades, X e Y, é dada pela expressão a seguir:

@@

@@
